Domain-agnostic Outlier Ranking Algorithms (DORA)---A Configurable Pipeline for Facilitating Outlier Detection in Scientific Datasets

### 2022-01-05 publications/frontiers/dora_outlier_detection_frontiers.tex
Convert manuscript from IAAI to Frontiers format. #79

\def\year{2021}\relax
%File: formatting-instructions-latex-2021.tex
%release 2021.2
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage{aaai21}  % DO NOT CHANGE THIS
\usepackage{times}  % DO NOT CHANGE THIS
\usepackage{helvet} % DO NOT CHANGE THIS
\usepackage{courier}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\frenchspacing  % DO NOT CHANGE THIS
\setlength{\pdfpagewidth}{8.5in}  % DO NOT CHANGE THIS
\setlength{\pdfpageheight}{11in}  % DO NOT CHANGE THIS
%\nocopyright
%PDF Info Is REQUIRED.
% For /Author, add all authors within the parentheses, separated by commas. No accents or commands.
% For /Title, add Title in Mixed Case. No accents or commands. Retain the parentheses.
\pdfinfo{
/Title (Domain-agnostic Outlier Ranking Algorithms (DORA)---A Configurable Pipeline for Facilitating Outlier Detection in Scientific Datasets)
/Author (AAAI Press Staff, Pater Patel Schneider, Sunil Issar, J. Scott Penberthy, George Ferguson, Hans Guesgen, Francisco Cruz, Marc Pujol-Gonzalez)
/TemplateVersion (2021.2)
} %Leave this
% /Title ()
% Put your actual complete title (no codes, scripts, shortcuts, or LaTeX commands) within the parentheses in mixed case
% Leave the space between \Title and the beginning parenthesis alone
% /Author ()
% Put your actual complete list of authors (no codes, scripts, shortcuts, or LaTeX commands) within the parentheses in mixed case.
% Each author should be only by a comma. If the name contains accents, remove them. If there are any LaTeX commands,
% remove them.

% DISALLOWED PACKAGES
% \usepackage{authblk} -- This package is specifically forbidden
% \usepackage{balance} -- This package is specifically forbidden
% \usepackage{color (if used in text)
% \usepackage{CJK} -- This package is specifically forbidden
% \usepackage{float} -- This package is specifically forbidden
% \usepackage{flushend} -- This package is specifically forbidden
% \usepackage{fontenc} -- This package is specifically forbidden
% \usepackage{fullpage} -- This package is specifically forbidden
% \usepackage{geometry} -- This package is specifically forbidden
% \usepackage{grffile} -- This package is specifically forbidden
% \usepackage{hyperref} -- This package is specifically forbidden
% \usepackage{navigator} -- This package is specifically forbidden
% (or any other package that embeds links such as navigator or hyperref)
% \indentfirst} -- This package is specifically forbidden
% \layout} -- This package is specifically forbidden
% \multicol} -- This package is specifically forbidden
% \nameref} -- This package is specifically forbidden
% \usepackage{savetrees} -- This package is specifically forbidden
% \usepackage{setspace} -- This package is specifically forbidden
% \usepackage{stfloats} -- This package is specifically forbidden
% \usepackage{tabu} -- This package is specifically forbidden
% \usepackage{titlesec} -- This package is specifically forbidden
% \usepackage{tocbibind} -- This package is specifically forbidden
% \usepackage{ulem} -- This package is specifically forbidden
% \usepackage{wrapfig} -- This package is specifically forbidden
% DISALLOWED COMMANDS
% \nocopyright -- Your paper will not be published if you use this command
% \addtolength -- This command may not be used
% \balance -- This command may not be used
% \baselinestretch -- Your paper will not be published if you use this command
% \clearpage -- No page breaks of any kind may be used for the final version of your paper
% \columnsep -- This command may not be used
% \newpage -- No page breaks of any kind may be used for the final version of your paper
% \pagebreak -- No page breaks of any kind may be used for the final version of your paperr
% \pagestyle -- This command may not be used
% \tiny -- This is not an acceptable font size.
% \vspace{- -- No negative value may be used in proximity of a caption, figure, table, section, subsection, subsubsection, or reference
% \vskip{- -- No negative value may be used to alter spacing above or below a caption, figure, table, section, subsection, subsubsection, or reference

\setcounter{secnumdepth}{0} %May be changed to 1 or 2 if section numbers are desired.

\usepackage{siunitx}
\usepackage{amsmath}
\usepackage[dvipsnames]{xcolor}
\newcommand{\todo}[1]{\textcolor{blue}{#1}}

% The file aaai21.sty is the style file for AAAI Press
% proceedings, working notes, and technical reports.
%

% Title

% Your title must be in mixed case, not sentence case.
% That means all verbs (including short verbs like be, is, using,and go),
% nouns, adverbs, adjectives should be capitalized, including both words in hyphenated terms, while
% articles, conjunctions, and prepositions are lower case unless they
% directly follow a colon or long dash
\iffalse
\title{Domain-agnostic Outlier Ranking Algorithms (DORA)---A Configurable Pipeline for Facilitating Outlier Detection in Scientific Datasets}
\author{
    %Authors
    % All authors must be in the same font size and format.
    Written by AAAI Press Staff\textsuperscript{\rm 1}\thanks{With help from the AAAI Publications Committee.}\\
    AAAI Style Contributions by Pater Patel Schneider,
    Sunil Issar,  \\
    J. Scott Penberthy,
    George Ferguson,
    Hans Guesgen,
    Francisco Cruz,
    Marc Pujol-Gonzalez
    \\
}
\affiliations{
    %Afiliations
    \textsuperscript{\rm 1}Association for the Advancement of Artificial Intelligence\\
    %If you have multiple authors and multiple affiliations
    % use superscripts in text and roman font to identify them.
    %For example,

    % Sunil Issar, \textsuperscript{\rm 2}
    % J. Scott Penberthy, \textsuperscript{\rm 3}
    % George Ferguson,\textsuperscript{\rm 4}
    % Hans Guesgen, \textsuperscript{\rm 5}.
    % Note that the comma should be placed BEFORE the superscript for optimum readability

    2275 East Bayshore Road, Suite 160\\
    Palo Alto, California 94303\\
    % email address must be in roman text type, not monospace or sans serif
    publications21@aaai.org

    % See more examples next
}
\fi
\iffalse
%Example, Single Author, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
\title{My Publication Title --- Single Author}
\author {
    % Author
    Author Name \\
}

\affiliations{
    Affiliation \\
    Affiliation Line 2 \\
    name@example.com
}
\fi

%Example, Multiple Authors, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
\title{Domain-agnostic Outlier Ranking Algorithms (DORA)---A Configurable Pipeline for Facilitating Outlier Detection in Scientific Datasets}
\author {
    % Authors
    Hannah Kerner,\textsuperscript{\rm 1}
    Umaa Rebbapragada, \textsuperscript{\rm 2}
    Kiri Wagstaff, \textsuperscript{\rm 2} 
    Steven Lu, \textsuperscript{\rm 2} 
    Bryce Dubayah, \textsuperscript{\rm 1} \\
    Eric Huff, \textsuperscript{\rm 2} 
    Jake Lee, \textsuperscript{\rm 2} 
    Vinay Raman, \textsuperscript{\rm 3}
    Sakshum Kulshrestha \textsuperscript{\rm 1}  \\
}
\affiliations {
    % Affiliations
    \textsuperscript{\rm 1} University of Maryland, College Park \\
    \textsuperscript{\rm 2} Jet Propulsion Laboratory, 
    									   California Institute of Technology \\
    \textsuperscript{\rm 3} Montgomery Blair High School \\
    hkerner@umd.edu, \{umaa.d.rebbapragada,
    										kiri.l.wagstaff,
    										you.lu,
    										eric.m.huff,
    										jake.h.lee\}@jpl.nasa.gov
}

\begin{document}

@@

\begin{abstract}
Automatic detection of outliers is
universally needed when working with scientific datasets, e.g., for cleaning 
datasets or flagging novel samples to guide instrument acquisition or
 scientific analysis.
We present Domain-agnostic Outlier Ranking Algorithms (DORA),
a configurable pipeline that facilitates application and evaluation of
outlier detection methods in a variety of domains. DORA allows users to 
configure experiments by specifying the location of their 
dataset(s), the input data type, feature extraction methods, and which 
algorithms should be applied. DORA supports image, raster, time series, 
or feature vector input data types and outlier detection methods that include
 Isolation Forest, DEMUD, PCA, RX detector, 
 Local RX, negative sampling, and probabilistic autoencoder. Each algorithm assigns an outlier score to each data sample. DORA
   provides results interpretation modules to help users 
   process the results, including sorting samples by outlier score, evaluating 
   the fraction of known outliers in $n$ selections, clustering groups of 
   similar outliers together, and web visualization. We demonstrated how DORA
   facilitates application, evaluation, and interpretation of outlier detection
   methods by performing
    experiments for three real-world datasets from Earth science, planetary
  science, and astrophysics, as well as one benchmark dataset 
  (MNIST/Fashion-MNIST). We found that no single algorithm
      performed best across all datasets, underscoring the need for a tool 
      that enables comparison of multiple algorithms.
\end{abstract}

\section{Introduction}
The ability to automatically detect out-of-distribution samples in large data 
sets is of interest for a wide variety of scientific domains. Depending on the
 application setting, this capability is also commonly referred to as anomaly
  detection, outlier detection, or novelty detection. More broadly, this is 
  referred to as out-of-distribution (OOD) detection. In general, the goal of 
  OOD detection systems is to identify samples that deviate from the majority
   of samples in a dataset in an unsupervised manner~\cite{pimentel2014review}. 
   In machine learning, these methods are commonly used for identifying 
   mislabeled or otherwise invalid samples in a dataset. When working with science datasets, OOD detection can be used for 
   cleaning datasets, e.g., flagging ground-truth labels with GPS or human
    entry error or identifying wrongly categorized objects in a catalog. 
    It could also be used for discovery, e.g., to flag novel samples in order 
    to guide instrument acquisition or scientific analysis. Another application
   is the detection of rare objects that are known to exist but the known
   examples are too few to create a large enough labeled dataset for 
   supervised classification algorithms.

@@

\section{Related Work}
Methods for outlier detection have been 
surveyed extensively and can be differentiated primarily based on how they 
score outliers
~\cite{pimentel2014review,chandola2009anomaly,markou2003novelty,
markou2003novelty2}.
%
Reconstruction-based methods construct a model of a dataset by learning 
a mapping between the input data and a lower-dimensional representation that
 minimizes the loss between the input and its reconstruction from the 
 low-dimensional representation~\citep{kerner2020comparison}. The 
 reconstruction error is used as the outlier score because samples
  that are unlike the data used to fit the model will be more poorly 
  reconstructed compared to inliers. Reconstruction-based methods
  include PCA~\cite{jablonski2015principal}, 
  autoencoders~\cite{richter2017safe}, and
  generative adversarial networks~\cite{akcay2018ganomaly}.
%
 Distance-based methods score outliers based on their distance from a
``background'' which can be defined in a variety of ways. For example,
the Reed-Xiaoli (RX) detector computes the 
Mahalanobis distance between each sample and the background dataset
defined by its mean and covariance matrix~\cite{reed1990adaptive}.
%
Sparsity-based methods such as isolation forest~\cite{liu2008isolation}
and local outlier factor~\cite{breunig2000lof}
 score outliers based on how isolated or sparse samples 
are in a given feature space. 
%
Probability distribution and density based methods estimate the underlying 
distribution or probability density of a dataset and score samples using
 likelihood. Examples include the probabilistic
autoencoder, which scores samples based on the log likelihood under the 
latent space distribution~\cite{bohm2020probabilistic}, Gaussian mixture 
Models, and kernel density estimators~\cite{chandola2009anomaly}. 
%
Other methods formulate outlier detection as supervised classification, 
usually with only one class constituted by 
known normal samples.
Such methods include one-class support vector 
machines~\cite{scholkopf1999support}
and negative sampling~\cite{sipple:neg-sampling20}.

@@

\paragraph{Data types.} We chose to implement data loaders for four
data types that are commonly used by the machine learning and domain science
communities: time series, feature vectors, images (grayscale or RGB), and 
$N$-band rasters. $N$-band rasters are images or grids in which every pixel
is associated with a location (e.g., latitude/longitude in degrees); 
most satellite data are distributed as rasters.
A data loader for each data type locates the data by the path(s)
defined in the configuration file and loads samples into a dictionary of numpy
 arrays indexed by the sample id. This \texttt{data\_dict} is then passed to each of the ranking algorithms.

\paragraph{Outlier ranking algorithms.} We implemented 7 unsupervised 
algorithms for scoring and 
ranking samples by outlierness. 
We chose these algorithms to include a diverse
set of approaches to scoring outliers since different algorithms may  
perform better for different datasets and use cases. 
%We describe each
%approach to scoring outliers and the associated methods below.

@@

\subparagraph{Reconstruction error.}
Principal component analysis (PCA) has been used for outlier detection by
scoring samples using the reconstruction error (here, the L2 norm)
between inputs and their inverse
 transformation from the principal subspace~\citep{kerner2020comparison}.
 DEMUD~\citep{wagstaff:demud13} differs from other
outlier ranking methods: instead of independently scoring all
observations, DEMUD incrementally identifies the most unusual
remaining item, then incorporates it into the model of ``known''
(non-outlier) observations before selecting the next most unusual
item.  DEMUD's goal is to identify diverse outliers and avoid
redundant selections.  Once an outlier is found, repeated
occurrences of that outlier are deprioritized.  Methods that score
samples independently maximize coverage of outliers, while DEMUD
maximizes fast discovery of distinct outlier types.

\subparagraph{Distance.}
 The 
Reed-Xiaoli (RX) detector is commonly used for anomaly detection in
multispectral and hyperspectral remote sensing. RX scores samples using
 the Mahalanobis
 distance between a sample and a background mean and covariance
 ~\citep{reed1990adaptive}. The local variant of RX (Local
 RX or LRX) can be used for image or raster data and scores each pixel in
  an image with respect to a window ``ring'' of pixels surrounding 
  it~\citep{molero2013analysis}. 
  LRX requires two parameters to define the size of the outer window 
  surrounding the pixel and the inner window
   around the target pixel to exclude from the background distribution.

\subparagraph{Sparsity.}
Isolation forests are a common sparsity-based method
that constructs many random binary trees from a 
dataset~\cite{liu2008isolation}. The outlier score for
a sample is quantified as the average distance from the root to the item’s 
leaf. Shorter distances are indicative of outliers because fewer splits were 
required to isolate the sample.

\subparagraph{Probability.}
The negative sampling algorithm is implemented by converting the unsupervised 
outlier ranking problem into a semi-supervised 
problem~\citep{sipple:neg-sampling20}. Negative (anomalous) 
examples are created by sampling from an expanded space defined by the minimum 
and maximum values of each dimension of the positive (normal) examples. The 
negative and positive examples are then used to train a random forest 
classifier. We use the posterior probabilities of the random forest classifier 
as outlier scores, which means that the observations with higher posterior 
probabilities are more likely to be outliers.
The probabilistic autoencoder is a generative model consisting of an
autoencoder trained to reconstruct input data which is interpreted
probabilistically after training using a normalizing flow on the autoencoder
latent space~\cite{bohm2020probabilistic}.
Samples are scored as outliers using the log likelihood in the latent
distribution, the autoencoder reconstruction error, or a combination of
both.

\paragraph{Results interpretation.}
Each of the outlier ranking algorithms returns an array containing the sample
index, outlier score, and selection index (index after sorting
by outlier score). DORA provides organization and visualization modules
intended to help users interpret and make decisions based on these outputs.
The simplest
 module saves a CSV of the samples sorted by their outlier score 
 (i.e., selection order). Clustering the top
  $N$ outlier selections can enable users to investigate the different types of 
 outliers that might be present in the dataset; this could be especially useful
 for separating outliers caused by noise or data artifacts vs. scientifically 
 interesting samples. We implemented the K-means and self-organizing maps 
 (SOMs) algorithms for clustering the top-N outliers. For use cases in which an
 evaluation dataset containing known outliers is available, we provide a module
 to assess how well algorithm selections correlate with known outliers. This is
 done by plotting the number of known outliers vs. number of selections made. 
We provide a module for plotting histograms of outlier scores to visualize the
 distribution of scores in the dataset (which may be, e.g., multimodal or 
 long-tailed). 
We developed a desktop application to easily visualize DORA results with the
Electron application framework and React frontend library. 
This enables fast and easy comparison of the results from
different methods. Figure \ref{fig:doravis} shows a screenshot of the
``Aggregate Table'' view, which displays all algorithm results 
side-by-side.

%We developed a desktop application to easily visualize DORA results with the
%Electron application framework and React frontend library. The application loads
%the DORA configuration file to locate the dataset and result CSVs. Then,
%it displays the ranked samples and their scores in a table sorted by their
%selection order. This allows for fast and easy comparison of the results of
%different methods. Figure \ref{fig:doravis} shows a screenshot of the
%``Aggregate Table'' view, which displays all results from different algorithms
%side-by-side.

@@

\subsubsection{Astrophysics: objects in Dark Energy Survey.}
Astronomical data sets are large and growing. Large modern optical
imaging surveys are producing catalogs of order $10^8$ stars
and galaxies, with dozens or hundreds of distinct measured features
for each entry. Discovery science becomes difficult at this data
volume: the scale is too large for expert human inspection, and separating real
astrophysical anomalies from non-astrophysical sources like detector artifacts or
satellite trails is a challenging problem for current methods.

@@

\subsubsection{Planetary: targets in Mars rover images.}
Mars exploration is fundamentally an exercise in discovery with the
goal of increasing our understanding of Mars's history, evolution,
composition, and currently active processes.  Outliers
identified in Mars observations can inspire new discoveries and inform
the choice of which areas merit follow-up or deeper
investigation~\cite{kerner2020comparison,wagstaff:rover-novelty20}.
We collected \num{72} images from the Navigation camera (Navcam) 
on the Mars Science Laboratory (MSL) rover
and employed Rockster~\cite{burl:rockster16}
(currently used by onboard rover software) to identify candidate rock
targets with an area of at least 100 pixels, yielding \num{1050}
targets.  We cropped out a \num{64}$\times$\num{64} pixel image
centered on each target.

@@

\subsubsection{Earth: satellite time series for ground observations.}
Many Earth science applications using satellite Earth observation (EO) data
 require ground-truth observations for identifying and modeling ground-
 identified objects in the satellite observations. These ground observations
 also serve as labels that are paired with satellite data inputs for machine 
 learning models. For example, a model trained to classify crop types in
 satellite observations requires ground-annotated labels of crop type. A widespread challenge for ground-
 annotated labels is that there are often points with erroneous location or 
 label information (e.g., due to GPS location error or human entry error) that
 need to be cleaned before the labels can be used for machine learning or
 other downstream uses. Automatically detecting these outliers could save 
 substantial time required for cleaning datasets and improve the performance of
 downstream analyses that rely on high-quality datasets.

@@

\subsubsection{Benchmark: MNIST and Fashion-MNIST.}
We used MNIST and F-MNIST to demonstrate DORA with a
traditional benchmark dataset. We used 60,000 images from 
F-MNIST as the training set and a test set of 1,000 images each from MNIST (outliers) and F-MNIST (inliers). 
 %Well-discriminating algorithms should assign 
%higher outlier scores to samples from MNIST than F-MNIST.

@@

\section{Results}
\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{figures/planetary_combined_plot.pdf}
    \caption{Number of known outliers ranked in top $N$ selections for 
    planetary rover dataset.}
    \label{fig:planetary_results}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This is just an example/guide for you to refer to when submitting manuscripts to Frontiers, it is not mandatory to use Frontiers .cls files nor frontiers.tex  %
% This will only generate the Manuscript, the final article will be typeset by Frontiers after acceptance.   
%                                              %
%                                                                                                                                                         %
% When submitting your files, remember to upload this *tex file, the pdf generated with it, the *bib file (if bibliography is not within the *tex) and all the figures.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Version 3.4 Generated 2018/06/15 %%%
%%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
%%% In http://www.ctan.org/ you can find the packages and how to install them, if necessary. %%%
%%%  NB logo1.jpg is required in the path in order to correctly compile front page header %%%

%\documentclass[utf8]{frontiersSCNS} % for Science, Engineering and Humanities and Social Sciences articles
%\documentclass[utf8]{frontiersHLTH} % for Health articles
\documentclass[utf8]{frontiersFPHY} % for Physics and Applied Mathematics and Statistics articles

\setcitestyle{square} % for Physics and Applied Mathematics and Statistics articles
\usepackage{url,hyperref,lineno,microtype,subcaption}
\usepackage[onehalfspacing]{setspace}

% packages added by authors
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage{amsmath}
\usepackage[dvipsnames]{xcolor}
\newcommand{\todo}[1]{\textcolor{blue}{#1}}

\linenumbers

% Leave a blank line between paragraphs instead of using \\

\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Kerner {et~al.}} %use et al only if is more than 1 author
\def\Authors{Hannah R. Kerner\,$^{1,*}$, Umaa Rebbapragada\,$^{2}$, 
Kiri L. Wagstaff\,$^{2}$, Steven Lu\,$^{2}$, Bryce Dubayah\,$^{1}$,
Eric Huff\,$^{2}$, Jake Lee\,$^{2}$, Vinay Raman\,$^{3}$, and
Sakshum Kulshrestha\,$^{1}$}
% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
\def\Address{$^{1}$Department of Geographical Sciences, University of Maryland, College Park, MD, USA \\
$^{2}$Jet Propulsion Laboratory, California Institute of Technology, Pasadena, California, USA \\
$^{3}$Montgomery Blair High School, Silver Spring, MD, USA}
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{Corresponding Author}

\def\corrEmail{hkerner@umd.edu}

\begin{document}
\onecolumn
\firstpage{1}

\title[Domain-agnostic Outlier Ranking Algorithms (DORA)]{Domain-agnostic Outlier Ranking Algorithms (DORA)---A Configurable Pipeline for Facilitating Outlier Detection in Scientific Datasets}

\author[\firstAuthorLast ]{\Authors} %This field will be automatically populated
\address{} %This field will be automatically populated
\correspondance{} %This field will be automatically populated

\extraAuth{}% If there are more than 1 corresponding author, comment this line and uncomment the next one.
%\extraAuth{corresponding Author2 \\ Laboratory X2, Institute X2, Department X2, Organization X2, Street X2, City X2 , State XX2 (only USA, Canada and Australia), Zip Code2, X2 Country X2, email2@uni2.edu}

@@

\begin{abstract}

Automatic detection of outliers is universally needed when working with 
scientific datasets, e.g., for cleaning datasets or flagging novel samples to 
guide instrument acquisition or scientific analysis.
We present Domain-agnostic Outlier Ranking Algorithms (DORA),
a configurable pipeline that facilitates application and evaluation of
outlier detection methods in a variety of domains. DORA allows users to 
configure experiments by specifying the location of their 
dataset(s), the input data type, feature extraction methods, and which 
algorithms should be applied. DORA supports image, raster, time series, 
or feature vector input data types and outlier detection methods that include 
Isolation Forest, DEMUD, PCA, RX detector, Local RX, negative sampling, 
and probabilistic autoencoder. Each algorithm assigns an outlier score to 
each data sample. DORA provides results interpretation modules to help users 
process the results, including sorting samples by outlier score, evaluating 
the fraction of known outliers in $n$ selections, clustering groups of 
similar outliers together, and web visualization. We demonstrated how DORA
facilitates application, evaluation, and interpretation of outlier detection
methods by performing experiments for three real-world datasets from Earth 
science, planetary science, and astrophysics, as well as one benchmark dataset 
(MNIST/Fashion-MNIST). We found that no single algorithm performed best across 
all datasets, underscoring the need for a tool that enables comparison of 
multiple algorithms.

\tiny
 \keyFont{ \section{Keywords:} astrophysics, planetary science, Earth science, outlier detection, novelty detection, out-of-distribution detection, anomaly detection} %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
\end{abstract}

\section{Introduction}

The ability to automatically detect out-of-distribution samples in large data 
sets is of interest for a wide variety of scientific domains. Depending on the
application setting, this capability is also commonly referred to as anomaly
detection, outlier detection, or novelty detection. More broadly, this is 
referred to as out-of-distribution (OOD) detection. In general, the goal of 
OOD detection systems is to identify samples that deviate from the majority
of samples in a dataset in an unsupervised manner~\cite{pimentel2014review}. 
In machine learning, these methods are commonly used for identifying 
mislabeled or otherwise invalid samples in a dataset. When working with 
science datasets, OOD detection can be used for 
cleaning datasets, e.g., flagging ground-truth labels with GPS or human
entry error or identifying wrongly categorized objects in a catalog. 
It could also be used for discovery, e.g., to flag novel samples in order 
to guide instrument acquisition or scientific analysis. Another application
is the detection of rare objects that are known to exist but the known
examples are too few to create a large enough labeled dataset for 
supervised classification algorithms.

@@

\section{Related Work}

Methods for outlier detection have been 
surveyed extensively and can be differentiated primarily based on how they 
score outliers
~\cite{pimentel2014review,chandola2009anomaly,markou2003novelty,
markou2003novelty2}.
%
Reconstruction-based methods construct a model of a dataset by learning 
a mapping between the input data and a lower-dimensional representation that
minimizes the loss between the input and its reconstruction from the 
low-dimensional representation~\citep{kerner2020comparison}. The 
reconstruction error is used as the outlier score because samples
that are unlike the data used to fit the model will be more poorly 
reconstructed compared to inliers. Reconstruction-based methods
include PCA~\cite{jablonski2015principal}, 
autoencoders~\cite{richter2017safe}, and
generative adversarial networks~\cite{akcay2018ganomaly}.
%
Distance-based methods score outliers based on their distance from a
``background'' which can be defined in a variety of ways. For example,
the Reed-Xiaoli (RX) detector computes the 
Mahalanobis distance between each sample and the background dataset
defined by its mean and covariance matrix~\cite{reed1990adaptive}.
%
Sparsity-based methods such as isolation forest~\cite{liu2008isolation}
and local outlier factor~\cite{breunig2000lof}
score outliers based on how isolated or sparse samples 
are in a given feature space. 
%
Probability distribution and density based methods estimate the underlying 
distribution or probability density of a dataset and score samples using
likelihood. Examples include the probabilistic
autoencoder, which scores samples based on the log likelihood under the 
latent space distribution~\cite{bohm2020probabilistic}, Gaussian mixture 
Models, and kernel density estimators~\cite{chandola2009anomaly}. 
%
Other methods formulate outlier detection as supervised classification, 
usually with only one class constituted by 
known normal samples.
Such methods include one-class support vector 
machines~\cite{scholkopf1999support}
and negative sampling~\cite{sipple:neg-sampling20}.

@@

\subsection{Data loaders} 
We chose to implement data loaders for four
data types that are commonly used by the machine learning and domain science
communities: time series, feature vectors, images (grayscale or RGB), and 
$N$-band rasters. $N$-band rasters are images or grids in which every pixel
is associated with a location (e.g., latitude/longitude in degrees); 
most satellite data are distributed as rasters.
A data loader for each data type locates the data by the path(s)
defined in the configuration file and loads samples into a dictionary of numpy
arrays indexed by the sample id. This \texttt{data\_dict} is then passed to 
each of the ranking algorithms.

\subsection{Outlier ranking algorithms} 
We implemented 7 unsupervised 
algorithms for scoring and 
ranking samples by outlierness. 
We chose these algorithms to include a diverse
set of approaches to scoring outliers since different algorithms may  
perform better for different datasets and use cases. 
We describe each
approach to scoring outliers and the associated methods below.

@@

\subsubsection{Reconstruction error}
Principal component analysis (PCA) has been used for outlier detection by
scoring samples using the reconstruction error (here, the L2 norm)
between inputs and their inverse
transformation from the principal subspace~\citep{kerner2020comparison}.
DEMUD~\citep{wagstaff:demud13} differs from other
outlier ranking methods: instead of independently scoring all
observations, DEMUD incrementally identifies the most unusual
remaining item, then incorporates it into the model of ``known''
(non-outlier) observations before selecting the next most unusual
item.  DEMUD's goal is to identify diverse outliers and avoid
redundant selections.  Once an outlier is found, repeated
occurrences of that outlier are deprioritized.  Methods that score
samples independently maximize coverage of outliers, while DEMUD
maximizes fast discovery of distinct outlier types.

\subsubsection{Distance}
The 
Reed-Xiaoli (RX) detector is commonly used for anomaly detection in
multispectral and hyperspectral remote sensing. RX scores samples using
the Mahalanobis
distance between a sample and a background mean and covariance
~\citep{reed1990adaptive}. The local variant of RX (Local
RX or LRX) can be used for image or raster data and scores each pixel in
an image with respect to a window ``ring'' of pixels surrounding 
it~\citep{molero2013analysis}. 
LRX requires two parameters to define the size of the outer window 
surrounding the pixel and the inner window
around the target pixel to exclude from the background distribution.

\subsubsection{Sparsity}
Isolation forests are a common sparsity-based method
that constructs many random binary trees from a 
dataset~\cite{liu2008isolation}. The outlier score for
a sample is quantified as the average distance from the root to the item’s 
leaf. Shorter distances are indicative of outliers because fewer splits were 
required to isolate the sample.

\subsubsection{Probability}
The negative sampling algorithm is implemented by converting the unsupervised 
outlier ranking problem into a semi-supervised 
problem~\citep{sipple:neg-sampling20}. Negative (anomalous) 
examples are created by sampling from an expanded space defined by the minimum 
and maximum values of each dimension of the positive (normal) examples. The 
negative and positive examples are then used to train a random forest 
classifier. We use the posterior probabilities of the random forest classifier 
as outlier scores, which means that the observations with higher posterior 
probabilities are more likely to be outliers.
The probabilistic autoencoder is a generative model consisting of an
autoencoder trained to reconstruct input data which is interpreted
probabilistically after training using a normalizing flow on the autoencoder
latent space~\cite{bohm2020probabilistic}.
Samples are scored as outliers using the log likelihood in the latent
distribution, the autoencoder reconstruction error, or a combination of
both.

\subsection{Results interpretation}
Each of the outlier ranking algorithms returns an array containing the sample
index, outlier score, and selection index (index after sorting
by outlier score). DORA provides organization and visualization modules
intended to help users interpret and make decisions based on these outputs.
The simplest
module saves a CSV of the samples sorted by their outlier score 
(i.e., selection order). Clustering the top
$N$ outlier selections can enable users to investigate the different types of 
outliers that might be present in the dataset; this could be especially useful
for separating outliers caused by noise or data artifacts vs. scientifically 
interesting samples. We implemented the K-means and self-organizing maps 
(SOMs) algorithms for clustering the top-N outliers. For use cases in which an
evaluation dataset containing known outliers is available, we provide a module
to assess how well algorithm selections correlate with known outliers. This is
done by plotting the number of known outliers vs. number of selections made. 
We provide a module for plotting histograms of outlier scores to visualize the
distribution of scores in the dataset (which may be, e.g., multimodal or 
long-tailed). 
We developed a desktop application to easily visualize DORA results with the
Electron application framework and React frontend library. 
This enables fast and easy comparison of the results from
different methods. 
%Figure \ref{fig:doravis} shows a screenshot of the
%``Aggregate Table'' view, which displays all algorithm results 
%side-by-side.
%
We developed a desktop application to easily visualize DORA results with the
Electron application framework and React frontend library. 
The application loads
the DORA configuration file to locate the dataset and result CSVs. Then,
it displays the ranked samples and their scores in a table sorted by their
selection order. This allows for fast and easy comparison of the results of
different methods. Figure \ref{fig:doravis} shows a screenshot of the
``Aggregate Table'' view, which displays all results from different algorithms
side-by-side.

@@

\subsection{Astrophysics: objects in Dark Energy Survey}
Astronomical data sets are large and growing. Large modern optical
imaging surveys are producing catalogs of order $10^8$ stars
and galaxies, with dozens or hundreds of distinct measured features
for each entry. Discovery science becomes difficult at this data
volume: the scale is too large for expert human inspection, and separating real
astrophysical anomalies from non-astrophysical sources like detector artifacts or
satellite trails is a challenging problem for current methods.

@@

\subsection{Planetary: targets in Mars rover images}
Mars exploration is fundamentally an exercise in discovery with the
goal of increasing our understanding of Mars's history, evolution,
composition, and currently active processes.  Outliers
identified in Mars observations can inspire new discoveries and inform
the choice of which areas merit follow-up or deeper
investigation~\cite{kerner2020comparison,wagstaff:rover-novelty20}.
We collected \num{72} images from the Navigation camera (Navcam) 
on the Mars Science Laboratory (MSL) rover
and employed Rockster~\cite{burl:rockster16}
(currently used by onboard rover software) to identify candidate rock
targets with an area of at least 100 pixels, yielding \num{1050}
targets.  We cropped out a \num{64}$\times$\num{64} pixel image
centered on each target.

@@

\subsection{Earth: satellite time series for ground observations}
Many Earth science applications using satellite Earth observation (EO) data
require ground-truth observations for identifying and modeling ground-
identified objects in the satellite observations. These ground observations
also serve as labels that are paired with satellite data inputs for machine 
learning models. For example, a model trained to classify crop types in
satellite observations requires ground-annotated labels of crop type. 
A widespread challenge for ground-
annotated labels is that there are often points with erroneous location or 
label information (e.g., due to GPS location error or human entry error) that
need to be cleaned before the labels can be used for machine learning or
other downstream uses. Automatically detecting these outliers could save 
substantial time required for cleaning datasets and improve the performance of
downstream analyses that rely on high-quality datasets.

@@

\subsection{Benchmark: MNIST and Fashion-MNIST}
We used MNIST and F-MNIST to demonstrate DORA with a
traditional benchmark dataset. We used 60,000 images from 
F-MNIST as the training set and a test set of 1,000 images 
each from MNIST (outliers) and F-MNIST (inliers). 
%Well-discriminating algorithms should assign 
%higher outlier scores to samples from MNIST than F-MNIST.

@@

\section{Results}

% TODO: Tried to make these a 2x2 panel of subfigures, but was getting
% errors using the subfigure package (Error: 'Not in outer par mode'). It seems
% this option may not be available for Frontiers articles 
% (see bottom of document). May combine them as a single figure manually.
